---
title: "Data Generation"
author: "Jihyun & Man"
date: "04/04/2020"
  html_document:
    toc: true
    toc_float: true
---

```{r, include = FALSE, message = F}
library(dplyr)
library(tidyr)
library(MASS)
library(mvtnorm)
library(metafor)
```
* Checking 

# Raw Data

```{r, eval = F}
set.seed(20200317)
k <-  100 # number of studies 
N <- 200  # N 

# --------------------------------
# generating sample sizes 
# --------------------------------
chisq <- rchisq(n = k, df = 3) 
samplesize <- round((N/2) * ((chisq - 3)/(sqrt(6)))  + N, digits = 0)

# --------------------------------
# population correlation matrix 
# --------------------------------
R <- matrix(cbind(1,   .18, .06,
                  .18, 1, .11,
                  .06, .11, 1), nrow = 3)

SD <- matrix(cbind(2, 0, 0, 
                   0, 2, 0,
                   0, 0, 7.67), nrow = 3)
# --------------------------------
# Sigma and Mean 
# --------------------------------
Sigma <- SD %*% R %*% t(SD) #covariance structure of observed variables
Mu_ctr <- c(10, 10, 50.73)

# -------------------------------------------------------------
# for loop - generating control group individual data 
# -------------------------------------------------------------
raw_dat <- list()


for (i in 1:k) {
  
  # Sample Size 
  N <- samplesize[i]
  
  # Control Group
  y_c <- rmvnorm(n = N, mean = Mu_ctr, sigma = Sigma) %>%
    data.frame() %>%
    mutate(group = "control",
           N = N)
  
  es <- 0.5 # this can be changed - we may use a vector of effect sizes (but here I used a single effect size)
  y_c_mean <- colMeans(y_c[,1:3])
  sd_c <- apply(y_c[,1:3], 2, sd)
  
  s_p1 <- sqrt(((N - 1)*sd_c[1]  + (N - 1)*sd_c[1])/( 2*N - 2))  # pooled sd for verbal fluency 
  s_p2 <- sqrt(((N - 1)*sd_c[2] + (N - 1)*sd_c[2])/( 2*N - 2))  # pooled sd for stroop color word
  s_p3 <- sqrt(((N - 1)*sd_c[3] + (N - 1)*sd_c[3])/( 2*N - 2))  # pooled sd for internalizing

   
  # Treatment Group  
  Mu_trt <- c(((es * s_p1) + y_c_mean[1]), ((es * s_p2) + y_c_mean[2]), ((es * s_p3) + y_c_mean[3]))

  y_t <- rmvnorm(n = N, mean = Mu_trt, sigma = Sigma) %>%
    data.frame() %>%
    mutate(group = "intervention",
           N = N)
  
  dat <- rbind(y_c, y_t)
  colnames(dat)[1:3] <- c("Verbal", "Stroop", "Internalzing")
  
  # binary variable 
  raw_dat[[i]] <- dat %>%
    mutate(Grp = ifelse(group == "control", 0, 1))
  
}
```

# Introducing missingness at variable-level
Kejin: please note that I made either "verbal" or "Stroop" missing, assuming that primary studies include either one of the two mediator variables. Let me know if you have any questions!

```{r, eval = F}
library(purrr)

# -------------------------------------------------------------
# a function to introduce missingness on either Verbal or Stroop variable
# -------------------------------------------------------------
intr_miss <- function(x) {
if(runif(1, 0, 1) <.4) { x[,c("Verbal")] <- NA}
  else{
    if(runif(1, 0, 1) <.4) {x[,c("Stroop")] <- NA}}
 return(x)
  }


raw_dat_miss <- purrr::map(raw_dat, intr_miss)
# please use --> raw_dat_miss

```

# Calculate Summary stats

```{r, warning=FALSE, message=FALSE}

sim_dat <- 
  bind_rows(raw_dat_miss, .id = "column_label")

sim_clean <- 
  sim_dat %>% 
  dplyr::rename(studyid = column_label, y1 = group, y2 = Verbal, y3 = Stroop, y4 = Internalzing) %>%
  dplyr::select(-N) %>% 
  mutate(studyid = as.numeric(studyid),
         y1 = ifelse(y1 == "control", "ctl", "trt")) %>% 
  group_by(studyid) %>%
  mutate(y2_dic = ifelse(y2 <= median(y2), 0, 1),
         y2_dic = factor(y2_dic)) %>% 
  ungroup()


sum_y3y4 <- 
  sim_clean %>%
  pivot_longer(cols = y3:y4,
               names_to = "variables",
               values_to = "values") %>% 
  group_by(studyid, y1, variables) %>% 
  summarise(n = n(),
            m = mean(values),
            sd = sd(values)) %>% 
  pivot_longer(cols = n:sd,
               names_to = "stat",
               values_to = "val")%>% 
  unite(stat, stat, y1, variables, sep = "_") %>% 
  pivot_wider(names_from = stat,
              values_from = val)%>% 
  mutate(n_ctl = n_ctl_y3,
         n_trt = n_trt_y3) %>% 
  dplyr::select(-c(n_ctl_y3, n_ctl_y4, n_trt_y3, n_trt_y4))


sum_y2 <- 
  sim_clean %>%
  group_by(studyid, y1, y2_dic) %>% 
  summarise(n = n()) %>% 
  mutate(y1_y2_dic = ifelse(is.na(y2_dic), NA, paste0(y1, "_y2_", y2_dic, sep = ""))) %>% #############
  ungroup() %>% 
  dplyr::select(studyid, y1_y2_dic, n) %>%
  filter(!is.na(y1_y2_dic)) %>% #############
  pivot_wider(names_from = y1_y2_dic,
              values_from = n)

  
sum_cor <- 
  sim_clean %>%
  group_by(studyid) %>% 
  mutate(cor23 = cor(y2, y3),
         cor24 = cor(y2, y4),
         cor34 = cor(y3, y4),
         cor23_v = (1 - cor23^2)^2 / (length(y2)-1),
         cor24_v = (1 - cor24^2)^2 / (length(y2)-1),
         cor34_v = (1 - cor34^2)^2 / (length(y3)-1)) %>%
  ungroup()  %>% 
  group_by(studyid, y1) %>% 
  mutate(cor23_y1 = cor(y2, y3),
         cor24_y1 = cor(y2, y4),
         cor34_y1 = cor(y3, y4)) %>% 
  ungroup() %>%
  dplyr::select(studyid, y1, starts_with("cor")) %>% 
  unique() %>% 
  pivot_longer(cols = cor23_y1:cor34_y1,
               names_to = "cor",
               values_to = "val") %>% 
  unite(cor, cor, y1) %>% 
  pivot_wider(names_from = cor,
              values_from = val)  %>%
  dplyr::rename_all(funs (c("studyid","cor23", "cor24", "cor34", 
                    "cor23_v", "cor24_v", "cor34_v", "cor23_ctl",
                    "cor24_ctl", "cor34_ctl", "cor23_trt", "cor24_trt", "cor34_trt")))

########200402
sum_y2y4 <- sim_clean %>%
  group_by(studyid, y2_dic)%>%
  summarise(n = n(),
            m = mean(y4),
            sd = sd(y4)) %>%
    pivot_wider(names_from = y2_dic,
              values_from = n:sd)%>% 
  dplyr::select(-c(n_NA, m_NA, sd_NA))%>%
  rename(n_0_y4 = n_0, n_1_y4 = n_1, m_0_y4=m_0, m_1_y4 = m_1, sd_0_y4 = sd_0, sd_1_y4 = sd_1)



summary_dat <- 
  sum_cor %>% 
  left_join(sum_y3y4, by = "studyid") %>% 
  left_join(sum_y2, by = "studyid") %>% 
  dplyr::select(studyid, starts_with("n_"), starts_with("cor"), starts_with("m"), 
                starts_with("sd"), starts_with("ctl_y2"), starts_with("trt_y2"))

summary_dat <- 
  summary_dat %>% 
  mutate(n_y2_0 = ctl_y2_0 + trt_y2_0,
         n_y2_1 = ctl_y2_1 + trt_y2_1)

###########update
summary_dat <- 
  summary_dat %>%
  left_join(sum_y2y4)
```


# Get correlation estimates
Between y1 (group) and y3 or y4 (interval) - Controlled experiment.

## Calculate d

### From sample statistics
```{r}
cal_d <- function(trt_m, trt_sd, trt_n, ctl_m, ctl_sd, ctl_n){
  d <- (trt_m - ctl_m)/sqrt(((trt_n-1)*trt_sd^2 + (ctl_n-1)*ctl_sd^2)/(trt_n + ctl_n -2))
  v_d <- (trt_n + ctl_n)/(trt_n*ctl_n) + d^2 /(2*(trt_n + ctl_n))
  cbind(d, v_d)
}

# y3
es_dat <- summary_dat %>% mutate(d_y3 = cal_d(m_trt_y3, sd_trt_y3, n_trt, m_ctl_y3, sd_ctl_y3, n_ctl)[,"d"], v_d_y3 = cal_d(m_trt_y3, sd_trt_y3, n_trt, m_ctl_y3, sd_ctl_y3, n_ctl)[,"v_d"])

# y4
es_dat <- es_dat %>% mutate(d_y4 = cal_d(m_trt_y4, sd_trt_y4, n_trt, m_ctl_y4, sd_ctl_y4, n_ctl)[,"d"], v_d_y4 = cal_d(m_trt_y4, sd_trt_y4, n_trt, m_ctl_y4, sd_ctl_y4, n_ctl)[,"v_d"])

```


### From t statistic to d
```{r}
## Calculate t stat
cal_t <- function(n_trt, m_trt, sd_trt, m_ctl, sd_ctl, n_ctl){
sp <- sqrt(((n_trt-1)*sd_trt^2 + (n_ctl-1)*sd_ctl^2)/(n_trt + n_ctl -2))
t <- (m_trt - m_ctl)/(sp*sqrt((1/n_ctl)+ (1/n_ctl)))
t
}

## Calculate t to d - we may not need this. 
cal_t_d <- function(t, trt_n, ctl_n){
  d <- t*sqrt(1/trt_n + 1/ctl_n) 
  d
}
```

```{r}
# Add into data
### y1-y3 and y1-y4
es_dat <- es_dat %>% 
  mutate(t_y3 = cal_t(n_trt, m_trt_y3, sd_trt_y3, m_ctl_y3, sd_ctl_y3, n_ctl),
         t_y4 = cal_t(n_trt, m_trt_y4, sd_trt_y4, m_ctl_y4, sd_ctl_y4, n_ctl))
```

## Calculate correlation: Exp design
Between y1 and continuous variables (y3, y4)
```{r}
cal_rce <- function(d, n1, n2) {
  a <- (n1+n2)^2/(n1*n2)
  r_ce <- d/sqrt(d^2 + a^2)
  r_ce
}

# large sample variance of r_ce
cal_v_rce <- function(d, n1, n2){
  a <- (n1+n2)^2/(n1*n2)
  v_rce <- (a^4/(d^2 + a^2)^3)*((n1+n2)/(n1*n2) + (d^2)/(2*(n1+n2)))
  v_rce
}

```


```{r}
es_dat <- es_dat %>%
  mutate(rce_y1y3 = cal_rce(d_y3, n_trt, n_ctl),
         v_rce_y1y3 = cal_v_rce(d_y3, n_trt, n_ctl),
         rce_y1y4 = cal_rce(d_y4, n_trt, n_ctl),
         v_rce_y1y4 = cal_v_rce(d_y4, n_trt, n_ctl))
```

## Calculate biserial correlation
Between y2 and y4
```{r}
# d to r_b
dtob <- function(d, n1, n2){
  h <- (n1 + n2 -2)/n1 + (n1+n2-2)/n2
  r_pb <- d/sqrt(d^2 + h)
  p <- n1/(n1+n2)
  c <- qnorm(p)
  r_b <- (sqrt(p*(1-p))/dnorm(c))*r_pb
return(r_b)
}
```

* Use n_y2_0 and n_y2_1 to calculate the biserial correlation between dichotomized y2 and y4.
* Sampstat -> d -> (r_pb -> r_b using `dtob`)

```{r}
# For this, we need to have d_y4 first. (from sample statistics using cal_d)
es_dat <- es_dat %>%
  mutate(rb_y4_d = dtob(d_y4, n_y2_0, n_y2_1))
```

### Sampling variance of r_b - Soper's approxiamte method.
```{r}
cal_v_rb <- function(r_b, n1, n2) {
p <- n1/(n1+n2)
c <- qnorm(p)
v_rb <- (1/(n1+n2-1))*((sqrt(p*(1-p))/dnorm(c) - r_b^2)^2)
return(v_rb)
}
```

```{r}
es_dat <- es_dat %>%
  mutate(v_rb_y4 = cal_v_rb(rb_y4_d, n_trt, n_ctl))
```


## y1 (group) & y2 (dichotomous)

* Tetrachoric correlation assumes underlying continuous distribution of two dichotomous variables.

### From contingency table - cell size

```{r}
# function from John Fox's polychor function
# This uses ML estimator. 

pcor <- function(x) {
  require(polycor)
  mat <- matrix(x, 2, 2)
  pc <- polychor(x = mat, ML = TRUE, std.err=TRUE)
  c(pcor = pc$rho, v_pcor = pc$var[1,1])
}

es_dat_y2 <- subset(es_dat, !is.na(ctl_y2_0)) #############

rho <- 
  apply(as.matrix(es_dat_y2[,grep("y2",names(es_dat_y2))]), 1, pcor) %>%  #############
  t() %>% as.data.frame() %>%
  mutate(studyid = es_dat_y2$studyid) %>%  #############
  dplyr::select(studyid, pcor, v_pcor)

# function from ?
calc_tet <- function(a00, a11, a10, a01) {
  or <- a00 * a11 / (a01 * a10)
  v_or <- 1/a00 + 1/a01 + 1/a10 + 1/a11
  delta <- 1 + sqrt(a00 * a11 / (a10 * a01))
  r_tet <- cos(pi / delta) # from Pearson, 1900
  n <- a00 + a11 + a10 + a01
  p0 <- (a00 + a01) / n
  p1 <- (a00 + a10) / n
  z0 <- qnorm(p0)
  z1 <- qnorm(p1)
  h0 <- dnorm(z0)
  h1 <- dnorm(z1)
  v_tet <- (p0 * (1 - p0) * p1 * (1 - p1)) / (n * (h0 * h1)^2)
  return(cbind(or, v_or, r_tet, v_tet))
}



# es_dat1 includes tetrachoric correlation, OR, v_or
es_dat1 <- 
  es_dat %>%
  left_join(rho, by = "studyid") %>% 
  mutate(or = calc_tet(ctl_y2_0, trt_y2_1, ctl_y2_1, trt_y2_0)[, "or"],
         v_or = calc_tet(ctl_y2_0, trt_y2_1, ctl_y2_1, trt_y2_0)[, "v_or"]) ###### Deleted unnecessary parts.
```

### OR and marginal Ns 
```{r}
#### If we have OR and marginal Ns for y2 (n0, n1) and marginals for y1 (n_ctl, n_trt) and Total N
calc_tet2 <- function(OR, n0, n_ctl, N){
  delta <- 1 + sqrt(OR)
  r_tet <- cos(pi / delta)
  
  p0 <- n0/N
  p1 <- n_ctl/N
  z0 <- qnorm(p0)
  z1 <- qnorm(p1)
  h0 <- dnorm(z0)
  h1 <- dnorm(z1)
  v_tet <- (p0 * (1 - p0) * p1 * (1 - p1)) / (N * (h0 * h1)^2)
  return(cbind(r_tet, v_tet))
}


```


# Introduce missingness at the scenario level
```{r}
set.seed(20200401)
scenario <- c("s2", "s3", "s4")
y3_scenario <- sample(scenario, 100, replace = TRUE)
y4_scenario <- sample(scenario, 100, replace = TRUE)

scenario2 <- c("CellSize", "OR")
y2_scenario <- sample(scenario2, 100, replace = TRUE)

scenario3 <- c("summary", "cor")
y2y4_sce <- sample(scenario3, 100, replace = TRUE)

es_dat_final <-
  es_dat1 %>% 
  mutate(y13_scenario = y3_scenario,
         y14_scenario = y4_scenario,
         y12_scenario = y2_scenario,
         y24_scenario = y2y4_sce)
```

## y13

### scenario 2

```{r}
cal_r_s2 <- function(trt_m, trt_sd, trt_n, ctl_m, ctl_sd, ctl_n){
  # d
  d <- (trt_m - ctl_m)/sqrt(((trt_n-1)*trt_sd^2 + (ctl_n-1)*ctl_sd^2)/(trt_n + ctl_n -2))
  
  # controlled experiment r
  a <- (trt_n+ctl_n)^2/(trt_n*ctl_n)
  r_ce <- d/sqrt(d^2 + a^2)
  
  # variance of r_ce
  v_rce <- (a^4/(d^2 + a^2)^3)*((trt_n+ctl_n)/(trt_n*ctl_n) + (d^2)/(2*(trt_n+ctl_n)))
  
  cbind(r_ce, v_rce)
}


# Data
es_mis_s2 <- 
  es_dat_final %>%
  filter(y13_scenario=="s2") %>%
mutate(final_r_13 = cal_r_s2(m_trt_y3, sd_trt_y3, n_trt, m_ctl_y3, sd_ctl_y3, n_ctl)[,"r_ce"],
       final_vr_13 = cal_r_s2(m_trt_y3, sd_trt_y3, n_trt, m_ctl_y3, sd_ctl_y3, n_ctl)[,"v_rce"]) %>%
  dplyr::select(studyid, m_trt_y3, sd_trt_y3, n_trt, m_ctl_y3, sd_ctl_y3, n_ctl, y13_scenario, final_r_13, final_vr_13)
```

### scenario 3
When we already have d. 
```{r}
cal_r_s3 <- function(d, ctl_n, trt_n){
  # controlled experiment r
  a <- (trt_n+ctl_n)^2/(trt_n*ctl_n)
  r_ce <- d/sqrt(d^2 + a^2)
  
  # variance of r_ce
  v_rce <- (a^4/(d^2 + a^2)^3)*((trt_n+ctl_n)/(trt_n*ctl_n) + (d^2)/(2*(trt_n+ctl_n)))
  
  cbind(r_ce, v_rce)
}


es_mis_s3 <- 
  es_dat_final %>%
  filter(y13_scenario=="s3") %>%
mutate(final_r_13 = cal_r_s3(d_y3, n_trt, n_ctl)[,"r_ce"],
       final_vr_13 = cal_r_s3(d_y3, n_trt, n_ctl)[,"v_rce"]) %>%
  dplyr::select(studyid,d_y3, n_trt, n_ctl, y13_scenario, final_r_13, final_vr_13)
```

### scenario 4
When we have t.
```{r}
cal_r_s4 <- function(t, ctl_n, trt_n){
  # calculate d from t
  d <- t*sqrt(1/trt_n + 1/ctl_n)
  
  # controlled experiment r
  a <- (trt_n+ctl_n)^2/(trt_n*ctl_n)
  r_ce <- d/sqrt(d^2 + a^2)
  
  # variance of r_ce
  v_rce <- (a^4/(d^2 + a^2)^3)*((trt_n+ctl_n)/(trt_n*ctl_n) + (d^2)/(2*(trt_n+ctl_n)))
  
  cbind(r_ce, v_rce)
}


es_mis_s4 <- 
  es_dat_final %>%
  filter(y13_scenario=="s4") %>%
mutate(final_r_13 = cal_r_s4(t_y3, n_trt, n_ctl)[,"r_ce"],
       final_vr_13 = cal_r_s4(t_y3, n_trt, n_ctl)[,"v_rce"]) %>%
  dplyr::select(studyid,t_y3, n_trt, n_ctl, y13_scenario, final_r_13, final_vr_13)
```

## y14

```{r}
# scenario 2
es_mis_s2_y4 <- 
  es_dat_final %>%
  filter(y14_scenario=="s2") %>%
mutate(final_r_14 = cal_r_s2(m_trt_y4, sd_trt_y4, n_trt, m_ctl_y4, sd_ctl_y4, n_ctl)[,"r_ce"],
       final_vr_14 = cal_r_s2(m_trt_y4, sd_trt_y4, n_trt, m_ctl_y4, sd_ctl_y4, n_ctl)[,"v_rce"]) %>%
  dplyr::select(studyid,m_trt_y4, sd_trt_y4, n_trt, m_ctl_y4, sd_ctl_y4, n_ctl, y14_scenario, final_r_14, final_vr_14)

# scenario 3
es_mis_s3_y4 <- 
  es_dat_final %>%
  filter(y14_scenario=="s3") %>%
mutate(final_r_14 = cal_r_s3(d_y4, n_trt, n_ctl)[,"r_ce"],
       final_vr_14 = cal_r_s3(d_y4, n_trt, n_ctl)[,"v_rce"]) %>%
  dplyr::select(studyid,d_y4, n_trt, n_ctl, y14_scenario, final_r_14, final_vr_14)

# scenario 4
es_mis_s4_y4 <- 
  es_dat_final %>%
  filter(y14_scenario=="s4") %>%
mutate(final_r_14 = cal_r_s4(t_y4, n_trt, n_ctl)[,"r_ce"],
       final_vr_14 = cal_r_s4(t_y4, n_trt, n_ctl)[,"v_rce"]) %>%
  dplyr::select(studyid,t_y4, n_trt, n_ctl,y14_scenario, final_r_14, final_vr_14)

```

## cor y1 y2
cellsize - pcor, v_pcor
```{r}
es_mis_cel_y12 <- es_dat_final %>%
  filter(y12_scenario=="CellSize")%>%
  mutate(final_r_12 = pcor,
         final_vr_12 = v_pcor)%>%
  dplyr::select(studyid,ctl_y2_0, trt_y2_1, ctl_y2_1, trt_y2_0, y12_scenario, final_r_12, final_vr_12)
```

or 
```{r}
es_mis_or_y12 <- es_dat_final %>%
  filter(y12_scenario=="OR")%>%
  mutate(final_r_12 = calc_tet2(or, n_y2_0, n_ctl, N = n_ctl + n_trt)[,"r_tet"],
         final_vr_12 = calc_tet2(or, n_y2_0, n_ctl, N = n_ctl + n_trt)[,"v_tet"])%>%
  dplyr::select(studyid, or,n_y2_0, n_ctl, n_trt, y12_scenario, final_r_12, final_vr_12)
```

## cor y2 y4
r
```{r}
es_mis_r_y24 <- es_dat_final %>%
  filter(y24_scenario=="cor") %>%
  mutate(final_r_24 = cor24,
         final_vr_24 = cor24_v) %>%
  dplyr::select(studyid, cor24, cor24_v, y24_scenario, final_r_24, final_vr_24) # no variance?
```

summary
```{r}
cal_r_y24 <- function(trt_m, trt_sd, trt_n, ctl_m, ctl_sd, ctl_n){
  # calculate d  
  d <- (trt_m - ctl_m)/sqrt(((trt_n-1)*trt_sd^2 + (ctl_n-1)*ctl_sd^2)/(trt_n + ctl_n -2))
  
  # calculate biserial
  h <- (trt_n + ctl_n -2)/trt_n + (trt_n+ctl_n-2)/ctl_n
  r_pb <- d/sqrt(d^2 + h)
  p <- trt_n/(trt_n+ctl_n)
  c <- qnorm(p)
  r_b <- (sqrt(p*(1-p))/dnorm(c))*r_pb
  
  return(r_b)
}


es_mis_y24 <- es_dat_final %>%
  filter(y24_scenario=="summary") %>%
  mutate(final_r_24 = cal_r_y24(m_1_y4, sd_1_y4, n_y2_1, m_0_y4, sd_0_y4, n_y2_0),
         final_vr_24 = cal_v_rb(final_r_24, n_y2_1,  n_y2_0)) %>%
  dplyr::select(studyid, cor24, y24_scenario, final_r_24, final_vr_24)
```



## Combine
```{r}
es_mis_y1y3 <- bind_rows(es_mis_s2, es_mis_s3, es_mis_s4)
es_mis_y1y4 <- bind_rows(es_mis_s2_y4, es_mis_s3_y4, es_mis_s4_y4)
es_mis_y1y2 <- bind_rows(es_mis_cel_y12, es_mis_or_y12)
es_mis_y2y4 <- bind_rows(es_mis_r_y24, es_mis_y24)

es_mis_dat <- es_mis_y1y3 %>%
  left_join(es_mis_y1y4, by="studyid") %>%
  left_join(es_mis_y1y2, by = "studyid") %>%
  left_join(es_mis_y2y4, by = "studyid")

save(es_mis_dat, file = "es_mis_dat.RData")
```

